\documentclass[a4paper,12pt,openright,twoside]{book}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[dvips]{graphicx}
\usepackage{makeidx}
\usepackage{moreverb}
\usepackage{multirow}
\usepackage{parskip}
\usepackage{longtable}
\usepackage{rotating}
\usepackage{array}
\usepackage{listings}
\usepackage[amssymb]{SIunits}
%\usepackage{hyperref}
\usepackage{color}
\usepackage[page,title]{appendix}
\usepackage[onehalfspacing]{setspace}
\newcommand{\mat}{\mathbf}
\usepackage[justification=centering, font=singlespacing]{caption}
% \oddsidemargin 0.72in \evensidemargin 0.12in
%\textwidth 6.15in \topmargin -0.60in \headheight -.0in \textheight 9.210in
\usepackage{vmargin}
\setpapersize{A4}
\setmarginsrb{40mm}{20mm}{20mm}{20mm}
{0pt}{0pt}{12pt}{30pt}
\pagestyle{plain}
%\newenvironment{refs}{
%\begin{list}{}{\leftmargin=0.5cm \topsep=0.06cm \itemindent=-0.5cm
%\listparindent=-0.5cm \parsep=0.035cm
%\itemsep=0.0cm}\item[]}{\end{list}}

%floats
%\renewcommand\floatpagefraction{.9}
%\renewcommand\topfraction{.9}
%\renewcommand\bottomfraction{.9}
%\renewcommand\textfraction{.1}   
%\setcounter{totalnumber}{50}
%\setcounter{topnumber}{50}
%\setcounter{bottomnumber}{50}

%\includeonly{Introduction}

\begin{document}
\begin{titlepage}
\vspace*{\fill}
\begin{center}
{\huge Derivation and Analysis of Malaria Parasite Clearance Times}\\[1cm]
by\\[1cm]
{\large Nicholas Robert George Shrine}\\[4cm]
Submitted in partial fulfilment of the requirements for the MSc in Statistics\\
Department of Probability \& Statistics\\[2cm]
September 2009
\end{center}
\vspace*{\fill}
\end{titlepage}
\clearpage{\pagestyle{empty}\cleardoublepage}
\frontmatter
Abstract of: Derivation and Analysis of Malaria Parasite Clearance Times\\[0.5cm]
Author: Nicholas Robert George Shrine\\[0.5cm]
Date: September 2009\\[0.1cm]
\setlength{\parindent}{0.4in}
\setlength{\parskip}{0.0in}

A clinical trial by GlaxoSmithKline was performed to compare the parasite clearance times for an antimalarial drug used alone and in combination with another drug. The primary endpoint is the time at which 90\% of the parasites have been cleared from the blood (PC90). The data consists of parasite counts per microlitre of blood before first dose of antimalarial and then at multiple times up to 48 hours from first dose for each of 43 subjects. There were 22 male and 21 female subjects, with 23 on the single-drug treatment and 20 on the combined-drug. Subjects were treated at two centres, with 24 and 19 at each respectively. 

Three methods were investigated for making interpolated estimates of the PC90 time between times at which parasite counts were taken: linear cubic polynomial regression, non-linear logistic regression and simple log-linear interpolation. The log-linear interpolation method was identified as the most suitable with problems with logistic regression previously applied to this data highlighted.

The estimated PC90 clearance times were analysed by 3-way ANOVA by centre, sex and treatment using appropriate variance stabilising transformations. There is evidence to reject the hypothesis of no interaction of sex and treatment ($p<0.05$). It is found that female subjects on the combined treatment have mean PC90 13.1 hours shorter (95\% confidence interval: 5.0, 22.7 hours) than on the single-drug treatment. There is no evidence to reject the hypothesis at the 5\% level that male subjects have the same mean PC90 on both treatments. The results of non-parametric resampling tests confirm the conclusions of the parametric analysis.

No evidence of a difference in mean PC50 between groups was found at the 5\% level. Good evidence to reject the hypothesis of mean PC99 times being equal between treatment groups was found ($p<0.0005$), with mean PC99 times for subjects on the combined-drug treatment being 12 hours shorter (95\% CI: 8, 16) than for subjects on the single drug.

The decrease in clearance times for combined-drug therapy is observed in other similar studies, but no other study reviewed finds a sex-treatment interaction effect.

Functional data analysis, used to model the speed of clearance, finds that the combined-drug treatment reaches its maximum rate of clearance 15 hours before the single-drug treatment.
\pagebreak
\begin{center}
{\large Acknowledgements}\\[1cm]
\end{center}
The author acknowledges GlaxoSmithKline for the provision of the data analysed in this dissertation.
%I would also like to thank my dissertation supervisor Professor John Biggins and all the members of staff involved with the MSc course at Sheffield. 

\pagebreak
\vspace*{\fill}
{\large The views and opinion contained in this dissertation shall not be construed or interpreted whether directly or indirectly to be the views or opinions of any of the officers or employees of GlaxoSmithKline Research and Development Limited or any of its affiliated companies forming part of the GlaxoSmithKline group of companies. Further, reliance on the information contained in this dissertation is at sole risk of the user. The information is provided ``as is" without any warranty or implied term of any kind, either express or implied, including but not limited to any implied warranties or implied terms as to quality, fitness for a particular purpose or non infringement. All such implied terms and warranties are hereby excluded.}
\vspace*{\fill}
\tableofcontents
%\listoffigures 

\mainmatter
\setlength{\parindent}{0.0in}
\setlength{\parskip}{0.1in}
\include{Introduction}
\include{Data}
\include{Derivation}
\include{Analysis}
\include{Alternative}
\include{Discussion}

\addappheadtotoc
\appendix
\appendixpage
\include{Appendices}

\backmatter
\bibliographystyle{wileyj}
\renewcommand{\bibname}{References}
{\singlespace\bibliography{dissertation}}
\end{document}